---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# mobsim

The goal of the package is to facilitate understanding of scale-dependent biodiversity changes.

The package includes functions to simulated species distributions in space with controlled abundance distributions as well as controlled intraspecific aggregation. For analysis there are functions for species rarefaction and accumulation curves, species-area relationships,  endemics-area relationships and th distance-decay of community similarity.

## Installation

You can install mobsim from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("MoBiodiv/mobsim", build_vignettes = TRUE)
```

## Getting help

You can get an overview of the available functions in `mobsim`:

```{r, eval = FALSE}
?mobsim
```

Or have a look at tutorials in the vignette:

```{r, eval = FALSE}
browseVignettes("mobsim")
```


## Example

Here is an example of how to simulate two communities, which just differ in their
spatial aggregation of species, but have the same species abundance distribution
and the same total number of individuals.

### simulation of communities

```{r example}
library(mobsim)
comm_rand <- sim_poisson_community(s_pool = 30, n_sim = 300)
comm_agg <- sim_thomas_community(s_pool = 30, n_sim = 300, sigma = 0.05, mother_points = 1)
```
```{r, fig.width=8, fig.height=4.5}
par(mfrow = c(1,2))
plot(comm_rand)
plot(comm_agg)
```

### Analysis of spatially-explicit community data

`mobsim` mobsim offer functions to analyse spatially-explicit community data.
For example the species-area relationship of a community can be easily evaluated.

```{r}
sar_rand <- divar(comm_rand)
sar_agg <- divar(comm_agg)
```




