---
title: "Analysis of biodiversity patterns"
author: "Felix May"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Analysis of biodiversity patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This introduction shows the list of biodiversity patterns that can be analysed
using `MoBspatial`. In this vignette the patterns are evaluated on a simulated
data set, but of cause the same patterns can be also derived from real data.

# Simulate and explore data

First, we generate community data by simulating 500 individuals from 30 species
with a cluster size of `sigma = 0.05`.

```{r}
library(MoBspatial)
sim_dat1 <- sim_thomas_community(s_pool = 30,  n_sim = 500, sigma = 0.05)
```

Then we explore the generated community object. In the plot each dot represents one individual and the color indicates the species
identity.

```{r, fig.width = 5, fig.height = 5}
plot(sim_dat1)
summary(sim_dat1)
str(sim_dat1)
```

# Non-spatial patterns

## Species abundance distribution

A fundamental non-spatial pattern of a community is the abundance distribution,
i.e. the distribution of commonness and rarity in a community. For the abundance
distribution there is no function in `MoBspatial`, but you easily get the 
abundance distribution by counting the number of individuals of each species using
the R standard function `table()`

```{r}
abund <- table(sim_dat1$census$species)
```

A standard plot in community ecology is the *rank-abundance plot* where the 
abundance of each species is plotted vs. its rank from highest to lowest abundance.
As a standard log-scaling is used for the abundance axis in this plot.

```{r, fig.width = 5, fig.height = 5}
plot(sort(as.numeric(abund), decreasing = T), type="b", log="y",
     xlab="Species rank", ylab="Species abundance")
```

Of course the abundance distribution can be also visualized as a histogram. 
By tradition for the binning of species abundances logarithms with base 2 are used,
following the suggestion of Preston (1948).
This means the first abundance class includes species with just one invididual,
the second class with two individuals, the third class with 3-4, the fourth with
5-8 etc.

This binning method can be easily implemented using the R package [untb](https://CRAN.R-project.org/package=untb) by
Robin Hankin. 

```{r, fig.width = 5, fig.height = 5}
library(untb)
abund_dist <- preston(count(abund))
abund_dist

barplot(height = as.numeric(abund_dist), names.arg = names(abund_dist),
        xlab = "Abundance class", ylab ="No. of species")
```

## Rarefaction curve

Another important biodiversity pattern is the *rarefaction-curve*, which estimates
how the number of observed species increases with sample size of individuals.
The rarefaction curve assumes that individuals are sampled randomly and
independently (Gotelli & Colwell 2001). The rarefaction curve only depends on 
the abundance distribution of species and therefore the function of MoBspatial only
requires species abundances as argument.

```{r, fig.width = 5, fig.height = 5}
rc1 <- rare_curve(abund)
str(rc1)
plot(species ~ n, data = rc1, type = "l", xlab = "No. of individuals sampled",
     ylab = " Expected no. of species")
```

# Spatial patterns

As indicated by the name of the package the main strengths of MoBspatial are the
simulation and analysis of *spatial* biodiversity patterns. In the following all
functions for spatial pattern evaluation included in MoBspatial are introduced.

## Species-accumulation curve

Closely related to the *rarefaction curve* is the *species-accumulation curve*
(Gotelli & Colwell 2001). In contrast to the rarefaction curve for the derivation
of the species-accumulation curve individuals are not sampled randomly, but starting
from a focal individual always the closets neighbour individual is sampled and the
number of encountered species is counted. The resultting curve is derived as the 
average of the curves for each focal individual. The species-accumulation curve
is influenced by the abundance distribution, but also by the spatial aggregation
or regularity of conspecific individuals (i.e. individuals of the same species).
Therefore the species-accumulation curves requires individuals positions in addition
to the abundances.

It is comprehensive to plot the rarefaction and the species-accumulation curve
together. The difference between the two curves indicates aggregation or regularity
of conspecifics.

```{r, fig.width = 5, fig.height = 5}
sac1 <- accum_curve(sim_dat1)
plot(1:length(rc1), rc1, type = "l", xlab = "No. of individuals sampled",
     ylab = " Expected no. of species")
lines(1:length(sac1), sac1, col = "red")
legend("bottomright", legend = c("Rarefaction","Accumulation"), col = 1:2, lwd = 2)
```

## Diversity-area relationships

The most well-known spatial biodiversity pattern is the *species-area relationship* 
(SAR). In MoBspatial the function `DiVAR` (diversity-area relationships) calculates
the species richness in randomly located subplots (quadrats or rectangles) of 
different sizes. However, the function `DiVAR` calculates additional indices,
including the Shannon and Simpson diversity indices for each subplot, as well
as the number of endemic species, which are the species that *only* occur within, 
but not outside the subplot. Documentation for the Shannon and Simpson diversity
indices can be found in the package [vegan](https://cran.r-project.org/package=vegan)
in the function `?diversity`. 

The endemics-area relationship (EAR) has been suggested as important tool to
investigate the consequences of habitat loss for biodiversity (He & Hubbell 2011,
Keil et al. 2015). For the evaluation of SAR and EAR using MoBspatial a vector
with subplot sizes as proportion of the total community size has to be defined

```{r}
subplot_size <- c(0.01, 0.05, 0.1, 0.2, 0.3, 0.4, 0.5, 
                   0.6, 0.7, 0.8, 0.9, 0.95, 0.99, 1)
divar1 <- divar(sim_dat1, prop_area = subplot_size)
head(divar1)
```
Plot SAR and EAR

```{r, fig.width = 5, fig.height = 5}
plot(mean_spec ~ prop_area, data = divar1, type = "b", ylim = c(0, 30),
     xlab = "Proportion of area", ylab = "No. of species")
lines(mean_end ~ prop_area, data = divar1, type = "b", col = 2)
legend("topleft",c("SAR","EAR"), col = 1:2, lwd = 2)
```

## Distance decay of community similarity

The last spatial pattern provided by MoBspatial is the *distance decay of
community similarity*, which quantifies how quickly the similarity in species
composition between two subplots decreases with the distance between two subplots.

The function `dist_decay` makes use of the function `vegdist` from the package
[vegan](https://cran.r-project.org/package=vegan). Therefore you can consult
`?vegdist` for a list of available similarity indices.

Here is a demonstration how the distance decay can be estimated and visualized

```{r,fig.width=5, fig.height=5}
dd <- dist_decay(sim_dat1, prop_area = 0.05, n_samples = 30, method = "jaccard")
head(dd)

plot(similarity ~ distance, data = dd, xlab = "Distance", ylab = "Similarity")
dd.loess <- loess(similarity ~ distance, data = dd)
pred.sim <- predict(dd.loess)
lines(dd$distance, pred.sim, lwd=2)
```


# References

1. F. W. Preston 1948. The Commonness, and Rarity, of Species. Ecology 29:254-283.

2. Cotelli & Colwell 2001. Quantifying biodiversity: procedures and pitfalls in 
the measurement and comparison of species richness. Ecology Letters 4: 379–391.

3. He & Hubbell 2001. Species-area relationships always overestimate extinction
rates from habitat loss. Nature 473:368–371

4. Keil et al. 2015. On the decline of biodiversity due to area loss. 
Nature communications 6.

5. Morlon et al. 2008. A general framework for the distance–decay of similarity in ecological communities.
Ecology Letters 9:904–917.
