---
title: "Introduction to mobsim"
author: "Felix May"
date: "`r Sys.Date()`"
output:
 rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Introduction to mobsim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Installing mobsim

The package `mobsim` is available on GitHub and you can freely download 
the code there.

[mobsim on GitHub](https://github.com/MoBiodiv/mobsim)

However, when you just want to use the package, it is more easy to install the 
package directly from GitHub using the package `devtools`.

## Prerequistes

`mobsim` includes efficient C++ code that is integrated with R using the 
package `Rcpp`. Therefore you have to install the latest version of Rcpp 
from CRAN before installing `mobsim`.

You’ll also need a working C++ compiler. To get it:

* On Windows, install [Rtools](https://cran.r-project.org/bin/windows/Rtools/Rtools34.exe)
* On Mac, install Xcode from the app store.
* On Linux, sudo apt-get install r-base-dev or similar.

##Installation

Here is the code to install `mobsim` from GitHub

```{r, eval = F}
library(devtools)
install_github("MoBiodiv/mobsim", build_vignettes = T)
```

```{r, message=FALSE}
library(mobsim)
```

## Getting help

```{r, eval = F}
?mobsim
browseVignettes("mobsim")
```

# An example application of mobsim

This vignettes shows on example application of `mobsim` to illustrate the concepts
and functionality incorporated in the package. The example deals with the detection
of biodiversity changes in a landscape. Specifically, we address the question
of how inference on biodiversity changes depends on the biodiversity measures
used as well as the spatial scale of sampling.

The key idea of `mobsim` is to use controlled simulations to assess these changes.
In this example, we assume that a hypothetical biodiversity driver, e.g. 
an invasive species, pollution, or climate change reduces only the total number
of individuals in a landscape, but does not change the species abundance distribution
and the spatial distribution of species.

In a first step, we simulate communities, which reflect thess assumptions. In a 
second step, we generate virtual data sets by sampling the simulated communities, and
in the third step we analyse these data in a similar way, as we would analyse real
field data.

## Simulating communities

We simulate two communities that have the same species richness (`s_pool`),
the same Poisson log-normal species abundance distribution (SAD)
(specified by`sad_type` and `sad_coef`) of the species pool, and the same intraspecific 
aggregation of species with cluster size parameter `sigma`. The two communities only differ
in the total number of individuals (`n_sim`).

```{r}
sim_n_high <- sim_thomas_community(s_pool = 200, n_sim = 20000, sad_type = "poilog",
                                   sad_coef = list("cv_abund" = 1), sigma = 0.02)

sim_n_low <- sim_thomas_community(s_pool = 200, n_sim = 10000, sad_type = "poilog",
                                   sad_coef = list("cv_abund" = 1), sigma = 0.02)
```

The function `sim_thomas_community` generates community objects that can be 
conveniently summarised and plotted. The community object includes the xy-coordinates
and species identities of all simulated individuals.

```{r}
summary(sim_n_high)
summary(sim_n_low)
```

```{r, fig.width=7.2, fig.height=4.1}
par(mfrow = c(1,2))
plot(sim_n_high)
plot(sim_n_low)
```

## Analysis of spatially-explicit community data

The package `mobsim` offers functions to generate important biodiversity patterns
from spatially-explicit community objects, for instance the well-known species-
area relationship (SAR). The SAR is generated by nested subsampling of the
community with samples of different sizes. To generate the SAR, the function
`divar` (diversity-area relationships) is used, which requires a vector
of sampling areas, specified as proportion of the total area. Here, we use
an approximatley log-scaled sampling area vector.

```{r}
area <- c(0.001,0.002,0.005,0.01,0.02,0.05,0.1,0.2,0.5,1.0)
sar_n_high <- divar(sim_n_high, prop_area = area)
sar_n_low <- divar(sim_n_low, prop_area = area)

names(sar_n_high)
```

For each sampling size, the function calculates mean and standard deviation
of several biodiversity indices. For the SAR, we plot the mean species richness
vs. sampling area.

```{r, fig.width=5, fig.height=5}
plot(m_species ~ prop_area, data = sar_n_high, type = "b", log = "xy", ylim = c(2,200),
     xlab = "Proportion of area sampled.", ylab = "No. of species",
     main = "Species-area relationship")
lines(m_species ~ prop_area, data = sar_n_low, type = "b", col = "red")
legend("bottomright",c("N high","N low"), col = 1:2, pch = 1)
```

We see that there are more species in the landscape with more individuals across
most scales. However, the curves converge as soon as 50% or more of the landscape
are sampled.

In contrast to the simulated communities, where we have full information about
all indivivuals in the landscape, in reality we only have data from local samples,
e.g. plots, traps, etc.. In the following it is illustrated, how `mobsim` can
be used to simulate the sampling process.

## Sampling spatially-explicit community data

The function `sample_quadrats` distributes sampling quadrats of user defined
number (`n_quadrats`) and size (`quadrat_area`) in a landscape and returns the
species abundancce in each square. The user can choose different spatial sampling
designs. Here, we use a random distribution of the sampling squares.

First, we use many small samples:

```{r, fig.width=7.2, fig.height=4.1}
par(mfrow = c(1,2))
samples_S_n_high <- sample_quadrats(sim_n_high, n_quadrats = 100,
                                        quadrat_area = 0.001, method = "random",
                                        avoid_overlap = T)
samples_S_n_low <- sample_quadrats(sim_n_low, n_quadrats = 100,
                                       quadrat_area = 0.001, method = "random",
                                       avoid_overlap = T)
```

Second, we use less, but larger samples, that in total cover the same area than
the many small samples.

```{r, fig.width=7.2, fig.height=4.1}
par(mfrow = c(1,2))
samples_L_n_high <- sample_quadrats(sim_n_high, n_quadrats = 10,
                                  quadrat_area = 0.01, method = "random",
                                  avoid_overlap = T)
samples_L_n_low <- sample_quadrats(sim_n_low, n_quadrats = 10,
                                  quadrat_area = 0.01, method = "random",
                                  avoid_overlap = T)
```

The function `sample_quadrats` return two objects: (i) a community matrix
with quadrats in rows, species in columns and the abundance of every species in
the respective cell, and (ii) a matrix with the positions of the sampling
squares in the landscapes.

```{r}
dim(samples_L_n_high$spec_dat)
head(samples_L_n_high$spec_dat)[,1:5]

dim(samples_L_n_high$xy_dat)
head(samples_L_n_high$xy_dat)
```


## Analysis of plot-based community data

The community matrix can now be analysed just as ecologist will analyse field data.
A software package that is perfectly suited for this and smoothly integrates
with mobsim is [vegan](https://CRAN.R-project.org/package=vegan). 

Here, we use `vegan` to calculate the species richness, the Shannon- and Simpson-
diversity indices for both landscapes and for both sampling scales.

### Small scale analysis

First, we analyse the small-scale samples:

```{r, message=F}
library(vegan)
S_n_high <- specnumber(samples_S_n_high$spec_dat)
S_n_low <- specnumber(samples_S_n_low$spec_dat)

Shannon_n_low <- diversity(samples_S_n_high$spec_dat, index = "shannon")
Shannon_n_high <- diversity(samples_S_n_low$spec_dat, index = "shannon")

Simpson_n_low <- diversity(samples_S_n_high$spec_dat, index = "simpson")
Simpson_n_high <- diversity(samples_S_n_low$spec_dat, index = "simpson")
```

The three biodiversity indices are combined into one dataframe and can then
be conveniently analysed using boxplots.

```{r}
div_dat_S <- data.frame(N = rep(c("N high","N low"), each = length(S_n_high)),
                        S = c(S_n_high, S_n_low),
                        Shannon = c(Shannon_n_high, Simpson_n_low),
                        Simpson = c(Simpson_n_high, Simpson_n_low))
```

```{r, fig.width=7.2, fig.height=3.1}
par(mfrow = c(1,3))
boxplot(S ~ N, data = div_dat_S, ylab = "Species richness")
boxplot(Shannon ~ N, data = div_dat_S, ylab = "Shannon diversity")
boxplot(Simpson ~ N, data = div_dat_S, ylab = "Simpson diversity")
```

We see that on average the species richness and the Shannon-diversity are reduced
in the landscape with low total abundance, while surprisingly Simpson-diversity
is slightly higher with less individuals.

Since the different diversity indices have different absolute values, we calculate
relative changes of the mean values as diversity changes effect size.

The relative change is defined as

$$  relEff = \frac{diversity(N = low) - diversity(N = high)}{diversity(N = high)} $$

Accordingly, a relative effect of -0.5 means that diversity is reduced by 50%, while
a positive value means diversity increases by the reduction of total abundance.
A value of zero indicates no change.

```{r}
mean_div_S <- aggregate(div_dat_S[,2:4], by = list(div_dat_S$N), FUN = mean)
relEff_S <- (mean_div_S[mean_div_S$Group.1 == "N low", 2:4] -  mean_div_S[mean_div_S$Group.1 == "N high", 2:4])/
             mean_div_S[mean_div_S$Group.1 == "N high", 2:4]
relEff_S
```



These results indicate that species richness and Shannon decrease, while
Simpson surprisingly increases due to the reduction of total abundance. Since
the Simpson-index mainly measures the evenness of species abundances rather than
the number of species, this indicates that at the small scale local abundance
distributions decrease in the number of species, but increase in their evenness.

### Large scale analysis

Now we repeat this analysis using the large samples:

```{r}
S_n_high <- specnumber(samples_L_n_high$spec_dat)
S_n_low <- specnumber(samples_L_n_low$spec_dat)

Shannon_n_low <- diversity(samples_L_n_high$spec_dat, index = "shannon")
Shannon_n_high <- diversity(samples_L_n_low$spec_dat, index = "shannon")

Simpson_n_low <- diversity(samples_L_n_high$spec_dat, index = "simpson")
Simpson_n_high <- diversity(samples_L_n_low$spec_dat, index = "simpson")
```


```{r}
div_dat_L <- data.frame(N = rep(c("N high","N low"), each = length(S_n_high)),
                        S = c(S_n_high, S_n_low),
                        Shannon = c(Shannon_n_high, Simpson_n_low),
                        Simpson = c(Simpson_n_high, Simpson_n_low))
```

```{r, fig.width=7.2, fig.height=3.1}
par(mfrow = c(1,3))
boxplot(S ~ N, data = div_dat_L, ylab = "Species richness")
boxplot(Shannon ~ N, data = div_dat_L, ylab = "Shannon diversity")
boxplot(Simpson ~ N, data = div_dat_L, ylab = "Simpson diversity")
```

```{r}
mean_div_L <- aggregate(div_dat_L[,2:4], by = list(div_dat_L$N), FUN = mean)
relEff_L <- (mean_div_L[mean_div_S$Group.1 == "N low", 2:4] -  mean_div_L[mean_div_L$Group.1 == "N high", 2:4])/
             mean_div_L[mean_div_S$Group.1 == "N high", 2:4]
```

### Compare results across scales

Finally, we compare the diversity effect sizes across the two scales.

```{r}
relEff_S
relEff_L
```
We see that there are differences, both among indices within the same scale,
as shown above, but also across scales. At the larger scale there is a stronger
negative effect of shannon diversity, while the positive change in Simpson
diversity disappears. 

