---
title: "Introduction to MoBspatial"
author: "Felix May"
date: "`r Sys.Date()`"
output:
 rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Introduction to MoBspatial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Installing MoBspatial

The package `MoBspatial` is available on GitHub and you can freely download 
the code there.

[MoBspatial on GitHub](https://github.com/MoBiodiv/MoBspatial)

However, when you just want to use the package it is more easy to install the 
package directly from GitHub using the package `devtools`.

## Prerequistes

`MoBspatial` includes efficient C++ code that is integrated with R using the 
package `Rcpp`. Therefore you have to install the latest version of Rcpp 
from CRAN with install.packages("Rcpp").

You’ll also need a working C++ compiler. To get it:

* On Windows, install [Rtools](https://cran.r-project.org/bin/windows/Rtools/Rtools34.exe)
* On Mac, install Xcode from the app store.
* On Linux, sudo apt-get install r-base-dev or similar.

When you are interested furthere in `Rcpp` a good introduction is available at
in the free online book *Advanced R* by Hadley Wickham
(http://adv-r.had.co.nz/Rcpp.html)

##Installation

```{r, eval = F}
library(devtools)
install_github("MoBiodiv/MoBspatial", build_vignettes = T)
```

```{r}
library(MoBspatial)
```

## Getting help

```{r, eval = F}
?MoBspatial
browseVignettes("MoBspatial")
```

# Simulating communities

Biodiversity in any sampling area depends on three components

1. The total number of individuals
2. The relative abundances of species
3. The spatial distribution of species

`MoBspatial` provides functions to simulate communities and control for all 
these components.

## Simulate abundances

* Function `SAD.lognorm()`
* Choose
    + Number of species `S.pool`
    + Number of individuals `N.local`
    + Evenness of species abundance described by coefficient of variation of relative
    abundances `cv.abund`. Note the `cv.abund` is *inversely* related to evenness.
    cv is defined as standard deviation / mean abundance.
    
```{r}
sim1 <- SAD.lognorm(S.pool = 100, N.local = 1000, cv.abund = 0.0)
sim2 <- SAD.lognorm(S.pool = 100, N.local = 1000, cv.abund = 1.0)
sim3 <- SAD.lognorm(S.pool = 100, N.local = 1000, cv.abund = 2.0)
```

```{r}
str(sim1)
```

With fixed local species richness

```{r}
sim1a <- SAD.lognorm(S.pool = 100, N.local = 1000, cv.abund = 0.0, fixS.local = T)
sim2a <- SAD.lognorm(S.pool = 100, N.local = 1000, cv.abund = 1.0, fixS.local = T)
sim3a <- SAD.lognorm(S.pool = 100, N.local = 1000, cv.abund = 2.0, fixS.local = T)
```

```{r}
str(sim1a)
```

### Plotting abundance distributions

#### Rank-abundance plots

```{r}
abund1 <- sort(sim1$abund, decreasing = T)
abund2 <- sort(sim2$abund, decreasing = T)
abund3 <- sort(sim3$abund, decreasing = T)
```

```{r, fig.width = 5, fig.height = 5}
plot(as.numeric(abund1), type="b", log="y",
     xlab="Species rank", ylab="Species abundance", ylim = c(1, 200))
lines(1:length(abund2), as.numeric(abund2), col = 2, type = "b")
lines(1:length(abund3), as.numeric(abund3), col = 3, type = "b")
```

**How would a perfectly even community look like?**

#### Species abundance distributions

**Normal histogram**

```{r, fig.width = 8, fig.height = 8}
par(mfrow = c(2,2))
hist(abund1, xlab = "Abundance", ylab = "No. of species")
hist(abund2, xlab = "Abundance", ylab = "No. of species")
hist(abund3, xlab = "Abundance", ylab = "No. of species")
```

**Preston binning method**

* First class: 1 individual
* Second class: 2 individuals
* Third class: 3-4 individuals
* Fourth class: 5-8 individuals

```{r}
library(untb)
sad1 <- preston(abund1, n = 9)
sad2 <- preston(abund2, n = 9)
sad3 <- preston(abund3, n = 9)

sad1
```

```{r, fig.width = 5, fig.height = 5}
barplot(height = as.numeric(sad2), names.arg = names(sad1),
        xlab = "Abundance class", ylab ="No. of species", col = 2)
```

```{r, fig.width = 5, fig.height = 5}

abund_mat <- rbind(sad1, sad2, sad3)

barplot(height = abund_mat, names.arg = names(sad1), beside = T,
        xlab = "Abundance class", ylab ="No. of species", col = 1:3)
```

## Simulate spatial distribution

### Adding spatial coordinates to an abundance distribution

#### Random positions

Spatial poisson process: 

* Constant density of individuals in space
* Random distribution

```{r}
abund1 <- SAD.lognorm(S.pool = 10, N.local = 100)$abund
com1 <- Sim.Poisson.Coords(abund.vec = abund1, xrange = c(0,1), yrange = c(0,1))
```

Explore the community object

```{r}
class(com1)
str(com1)
summary(com1)
```

```{r, fig.width = 5, fig.height = 5}
plot(com1)
```

#### Aggregated positions

* Thomas process
   + Poisson distribution of mother points
   + Clustering of offspring points around mother points with cluster size `sigma`
   and number of offspring points

```{r}
com2 <- Sim.Thomas.Coords(abund.vec = abund1, sigma = 0.02)
```


```{r, fig.width = 5, fig.height = 5}
plot(com2)
```

**Change cluster radius**

```{r, fig.width = 10, fig.height = 5}
com2a <- Sim.Thomas.Coords(abund.vec = abund1, sigma = 0.05)
par(mfrow = c(1,2))
plot(com2)
plot(com2a)
```

**Change number of mother points**

```{r, fig.width = 10, fig.height = 5}
com2b <- Sim.Thomas.Coords(abund.vec = abund1, sigma = 0.02, mother.points = 1)
par(mfrow = c(1,2))
plot(com2)
plot(com2b)
```

**Change number of points per cluster**

```{r, fig.width = 10, fig.height = 5}
com2c <- Sim.Thomas.Coords(abund.vec = abund1, sigma = 0.02, cluster.points = 10)
par(mfrow = c(1,2))
plot(com2)
plot(com2c)
```

## Simulate abundances and distributions at the same time

### Poisson process

```{r}
com3 <- Sim.Poisson.Community(S.pool = 100, N.local = 500)
```

### Thomas process

```{r}
com4 <- Sim.Thomas.Community(S.pool = 100, N.local = 500, sigma = 0.05)
```

```{r, fig.width = 10, fig.height = 5}
par(mfrow = c(1,2))
plot(com3)
plot(com4)
```

