---
title: "Sampling communities with mobsim"
author: "Felix May"
date: "`r Sys.Date()`"
output:
 rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Sampling communities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 5)
```

In addition to the simulation and analysis of spatially-explicit communities,
`mobsim` provides a function to generate samples from simulated or observed communities. 
The combination of simulated data AND simulated sampling is a powerful approach
to test the validity and power of empirical approaches.

# Random sampling

Here, we simulate a community and then generate samples with the function
`sample_quadrats`. By default `sample_quadrats` distributes a user-defined
number of quadrats with user-defined size in the landscape and provides the
number of individuals for each species in each quadrat.

The function returns two dataframes. The first includes the abundance of every 
species in every sampling quadrat and the second the positions of the lower left
corners of the quadrats.

The community matrix of samples by species can then be analysed using additional 
software. For instance the R package [vegan](https://CRAN.R-project.org/package=vegan)
is perfectly suited for the analysis of community data. See the vignette **Introduction
to `mobsim`** for a worked example.

```{r}
library(mobsim)
sim_com1 <- sim_poisson_community(s_pool = 100, n_sim = 20000)
```

```{r, fig.width=4.4, fig.height=5}
sample1 <- sample_quadrats(sim_com1)

head(sample1$spec_dat[,1:6])
head(sample1$xy_dat)
```

In `sample_quadrats()` there is an option to exclude overlapping quadrats from the 
random sampling design, which is shown here in two examples with different numbers and sizes
of the quadrats.

```{r, fig.width=4.4, fig.height=5}
sample2 <- sample_quadrats(sim_com1, n_quadrats= 2, quadrat_area = 0.1,
                           avoid_overlap = T)
```

```{r, fig.width=4.4, fig.height=5}
sample3 <- sample_quadrats(sim_com1, n_quadrats= 20, quadrat_area = 0.001,
                           avoid_overlap = T)
```

# Transect sampling

In addition to random designs also transects can be sampled-. This requires
specifying a position for the lower left quadrat as well as x and y distances 
between neighbouring quadrats.

```{r, fig.width=4.4, fig.height=5}
sample4 <- sample_quadrats(sim_com1, n_quadrats= 10, quadrat_area = 0.005,
                           method = "transect", x0 = 0, y0 = 0.5, delta_x = 0.1,
                           delta_y = 0)

sample5 <- sample_quadrats(sim_com1, n_quadrats= 10, quadrat_area = 0.005,
                           method = "transect", x0 = 0, y0 = 0, delta_x = 0.1,
                           delta_y = 0.1)
```
                      
# Grid sampling

Finally, sampling quadrats can be arranged in a regular lattice. For this design
users have to choose distances among the quadrats in x and y dimension as shown
in the example.

```{r, fig.width=4.4, fig.height=5}
sample6 <- sample_quadrats(sim_com1, n_quadrats= 25, quadrat_area = 0.005,
                           method = "grid", x0 = 0, y0 = 0, delta_x = 0.1,
                           delta_y = 0.1)

sample7 <- sample_quadrats(sim_com1, n_quadrats= 25, quadrat_area = 0.005,
                           method = "grid", x0 = 0.05, y0 = 0.05, delta_x = 0.2,
                           delta_y = 0.2)
```

By default, `sample_quadrats()` plots the chosen design. However, the plotting
can be also deactivated for more efficient computations:

```{r}
sample7a <- sample_quadrats(sim_com1, n_quadrats= 25, quadrat_area = 0.005,
                           method = "grid", x0 = 0.05, y0 = 0.05, delta_x = 0.2,
                           delta_y = 0.2, plot = F)
head(sample7a$spec_dat[,1:10])
```


