---
title: "Simulation communities with mobsim"
author: "Felix May"
date: "`r Sys.Date()`"
output:
 rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Simulating communities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Biodiversity components

Biodiversity in any sampling area depends on three components:

1. The total number of individuals (total abundance)
2. The relative abundances of species
3. The spatial distribution of species

`mobsim` provides functions to simulate communities and control for all 
these components. This vignette first shows how to simulate non-spatial
species abundance distributions (SADs) and second, how to simulate spatially-
explicit community data with `mobsim`.

# Simulating species abundance distributions

For this purpose `mobsim` provides the function `sim_sad`, which is a wrapper
around the function `rsad` from the package [sads](https://CRAN.R-project.org/package=sads).
In contrast to `rsad`, `sim_sad` allows simulating communities with user defined
total number of individuals.

Here is an example for the simulation of an SAD using a log-normal model.

```{r}
library(mobsim)

abund1 <- sim_sad(s_pool = 100, n_sim = 1000, sad_type = "lnorm",
                  sad_coef = list("meanlog" = 5, "sdlog" = 0.5))
head(abund1)
summary(abund1)
```

`sim_sad` first simulates a *relative* species abundance distribution according
to the user defined model and then samples the requested number of individuals 
from this relative abundance distribution. Because of the use of a relative abundance
distribution as intermediate step, in the log-normal model the mean abundance 
is defined by the simulated number of individuals (`n_sim`) divided by the number 
of species (`s_pool`). Therefore, for the log-normal model `sim_sad` offers 
also a simpler parameterization that just specifies the coefficient of 
variation (`cv`) of the log-normal SAD.

```{r}
abund2 <- sim_sad(s_pool = 100, n_sim = 1000, sad_type = "lnorm",
                  sad_coef = list("cv_abund" = 2))
summary(abund2)
```

Obviously, the simulated community includes less species than the user defined 
value of `s_pool = 100`. This is a consequence of the stochastic sampling from the
relative abundance distribution. When some species have very low relative abundances,
they might not be sampled into the simulated community. However, `sim_sad` offers 
the option `fix_s_sim = TRUE`, which results in the user-defined value of species 
in the simulation. This is implemented by adding very rare species to the community,
while removing individuals from the common species. Please not that the constraint
`fix_s_sim = TRUE` might result in deviations from the underlying theoretical
SAD model.


```{r}
abund2a <- sim_sad(s_pool = 100, n_sim = 1000, sad_type = "lnorm",
                  sad_coef = list("cv_abund" = 2), fix_s_sim = T)
summary(abund2a)
```

The function `sim_sad` inherits all SAD models provided by `sads::rsad`.
For a complete list see `?sim_sad`. Here, we show an example of how to simulate
a log-series SAD. It has to be noted that for some SAD models the species richness
is not a direct parameter, but emerges from the other parameters. This is also true
for the log-series model. Therefore, the parameter `s_pool` is set to `NULL`.

```{r}
abund3 <- sim_sad(s_pool = NULL, n_sim = 10000, sad_type = "ls",
                  sad_coef = list("N" = 1e5, "alpha" = 20))
head(abund3)
```

Of course the simulated number of species can be easily evaluated

```{r}
length(abund3)
summary(abund3)
```

# Simulate spatial distributions

With `mobsim` random and aggregated species distributions can be simulated.
This can be done in two ways. Either, simulated coordinates of individuals 
can be added to an observed or simulated species abundance distributions, or
species abundance and coordinates can be simulated at the same time.

## Random distributions

In spatial statistics for point patterns a random distribution of points in a
given area is called Poisson process. Accordingly, th function to add random 
coordinates to an existing species abundance distribution is called `sim_poisson_coords`.
Here is an example of its application.


```{r}
abund1 <- c(20,10,10,5,5)
comm1 <- sim_poisson_coords(abund_vec = abund1, xrange = c(0,1), yrange = c(0,1))
```

The community object includes x and y coordinates, as well as the species
identity for every individual in the community. `mobsim` offers functions for
exploring and plotting the community object produces by `sim_poisson_coords`.


```{r, fig.width = 3.5, fig.height = 4}
class(comm1)
summary(comm1)
plot(comm1)
```

As mentioned above, abundances and (random) spatial distributions can be also
simulated at the same time using `sim_poisson_community`, which essentially calls
`sim_sad` and `sim_poisson_coords` consecutively.

```{r, fig.width = 3.5, fig.height = 4}
comm2 <- sim_poisson_community(s_pool = 20, n_sim = 200,
                               sad_type = "lnorm")
plot(comm2)
```


## Aggregated positions

Aggregated, or clustered species distributions are simulated based on the
Thomas process, also known as Poisson cluster process, in `mobsim`. For each
species the Thomas process first distributes a given number of mother points
in the landscape. Then, offspring points are distributed around the mother points
according to a bivariate Gaussion distance kernel, where the average distance
between mother and offspring points is called `sigma`. The offspring points
constitute the final distribution of the species. 

By variations in the size of clusters (`sigma`), the number of 
clusters (`mother_points`), and the mean number of individuals per cluster 
(`cluster_points`) the Thomas process can generate a large range of different
species distributions with intraspecific aggregation.

It is important to note the each species distribution is simulated independently
from the other species. That means the Thomas process in `mobsim` cannot be
used to simulate spatial dependence between different species, i.e. interspecific
aggregation or segregation.


Here is one example for a community with intraspecific aggregation:

```{r}
comm3 <- sim_thomas_coords(abund_vec = abund1, sigma = 0.02)
plot(comm3)
```

First, we change the size of the clusters

```{r, fig.width = 8, fig.height = 4}
comm3a <- sim_thomas_coords(abund_vec = abund1, sigma = 0.05)
par(mfrow = c(1,2))
plot(comm3)
plot(comm3a)
```

**Change number of mother points**

```{r, fig.width = 10, fig.height = 5}
com2b <- sim_thomas_coords(abund_vec = abund1, sigma = 0.02, mother_points = 1)
par(mfrow = c(1,2))
plot(com2)
plot(com2b)
```

**Change number of points per cluster**

```{r, fig.width = 10, fig.height = 5}
com2c <- sim_thomas_coords(abund_vec = abund1, sigma = 0.02, cluster_points = 10)
par(mfrow = c(1,2))
plot(com2)
plot(com2c)
```

## Simulate abundances and distributions at the same time

### Poisson process

```{r}
com3 <- sim_poisson_community(s_pool = 100, n_sim = 500)
```

### Thomas process

```{r}
com4 <- sim_thomas_community(s_pool = 100, n_sim = 500, sigma = 0.05)
```

```{r, fig.width = 10, fig.height = 5}
par(mfrow = c(1,2))
plot(com3)
plot(com4)
```





<!-- ```{r} -->
<!-- abund1 <-  -->
<!-- abund2 <- sim_sad(s_pool = 100, n_sim = 1000, sad_type = "lnorm", -->
<!--                 sad_coef = list("cv_abund" = 1.0)) -->
<!-- abund3 <- sim_sad(s_pool = 100, n_sim = 1000, sad_type = "lnorm", -->
<!--                 sad_coef = list("cv_abund" = 2.0)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- str(abund1) -->
<!-- ``` -->

<!-- With fixed local species richness -->

<!-- ```{r} -->
<!-- abund1a <- sim_sad(s_pool = 100, n_sim = 1000, sad_type = "lnorm", -->
<!--                 sad_coef = list("meanlog" = 2, "sdlog" = 0.5), fix_s_sim = T) -->
<!-- abund2a <- sim_sad(s_pool = 100, n_sim = 1000, sad_type = "lnorm", -->
<!--                 sad_coef = list("meanlog" = 2, "sdlog" = 1.0), fix_s_sim = T) -->
<!-- abund3a <- sim_sad(s_pool = 100, n_sim = 1000, sad_type = "lnorm", -->
<!--                 sad_coef = list("meanlog" = 2, "sdlog" = 2.0), fix_s_sim = T) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- str(abund1a) -->
<!-- ``` -->

<!-- ### Plotting abundance distributions -->

<!-- #### Rank-abundance plots -->

<!-- ```{r, fig.width = 5, fig.height = 5} -->
<!-- plot(as.numeric(abund1), type="b", log="y", -->
<!--      xlab="Species rank", ylab="Species abundance", ylim = c(1, 200)) -->
<!-- lines(1:length(abund2), as.numeric(abund2), col = 2, type = "b") -->
<!-- lines(1:length(abund3), as.numeric(abund3), col = 3, type = "b") -->
<!-- ``` -->

<!-- **How would a perfectly even community look like?** -->

<!-- #### Species abundance distributions -->

<!-- **Normal histogram** -->

<!-- ```{r, fig.width = 8, fig.height = 8} -->
<!-- par(mfrow = c(2,2)) -->
<!-- hist(abund1, xlab = "Abundance", ylab = "No. of species") -->
<!-- hist(abund2, xlab = "Abundance", ylab = "No. of species") -->
<!-- hist(abund3, xlab = "Abundance", ylab = "No. of species") -->
<!-- ``` -->

<!-- **Preston binning method** -->

<!-- * First class: 1 individual -->
<!-- * Second class: 2 individuals -->
<!-- * Third class: 3-4 individuals -->
<!-- * Fourth class: 5-8 individuals -->

<!-- ```{r} -->
<!-- oct_max <- ceiling(log2(max(c(abund1, abund2, abund3)))) -->
<!-- sad1 <- sads::octav(as.numeric(abund1), oct = 0:oct_max) -->
<!-- sad2 <- sads::octav(as.numeric(abund2), oct = 0:oct_max) -->
<!-- sad3 <- sads::octav(as.numeric(abund3), oct = 0:oct_max) -->

<!-- sad1 -->
<!-- ``` -->

<!-- ```{r, fig.width = 5, fig.height = 5} -->
<!-- class_names <- c("1","2","3-4","5-8","9-16","17-32","33-64","64-128","128-256",">256") -->
<!-- class_names <- class_names[1:nrow(sad1)] -->
<!-- barplot(height = sad2$Freq, names.arg = class_names, -->
<!--         xlab = "Abundance class", ylab ="No. of species", col = 2) -->
<!-- ``` -->

<!-- ```{r, fig.width = 5, fig.height = 5} -->

<!-- abund_mat <- rbind(sad1$Freq, sad2$Freq, sad3$Freq) -->

<!-- barplot(height = abund_mat, names.arg = class_names, beside = T, -->
<!--         xlab = "Abundance class", ylab ="No. of species", col = 1:3) -->
<!-- ``` -->

